Topological order in the quantum dimer-pentamer model

### 2016-04-19 qdpm_paper.tex
Worked on Background section

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

@@

\begin{figure}[b!]
    \centering
    \includegraphics[width=0.5\columnwidth]{QDPM_ex_config.pdf}
    \caption{Example of an allowed configuration in the quantum dimer-pentamer model.}
    \label{fig:QDPMex}
\end{figure}

@@

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.8\linewidth]{example_local_constraints.pdf}
    \caption{Clockwise from the upper left panel we show an example of a QDM with constraints: one
    dimer, two dimers, an even number of dimers, and and odd number of dimers touching each vertex.}
    \label{fig:example_local_constraints}
\end{figure}

Here we will provide background about generalized quantum dimer models on the square lattice. We consider systems with Ising-like dimer degrees of freedom on each link, such that each link has either $0$ or $1$ dimers. The Hilbert space consists of configurations of dimers on the square lattice that satisfy a local constraint at each vertex; the simplest constraint is that of the hard-core fully-packed QDM where a single dimer must touch each vertex. We write this constraint as $n_v = 1$, where $n_v$ is the number of dimers that touch a vertex. We can distinguish between number constraints of the form $n_v = n_0$ and parity constraints of the form $n_v\mod 2 = 0,1$. The unique, non-trivial constraints of this form are the number constraints $n_v=1,2$ and the parity constraints $n_v\mod2 =0,1$ and  $n_v\mod2 =1$ (see \figref{fig:example_local_constraints}).

The Hamiltonian of a GQDM is of the form
\begin{equation}
H_{\rm{GQDM}} = -t \hat{T} + v \hat{V}
\end{equation}
where $\hat{T}$ is an off-diagonal kinetic term that preserves the constraint, $\hat{V}$ is a diagonal potential energy term, and $v$ and $t$ parametrize the strength of each term. In the simplest case, $\hat{T}$ involves the minimum non-trivial dynamics that preserve the constraint. For example in the QDM $\hat{T}$ is the sum of all possible ``plaquette flip" terms acting on plaquettes with parallel dimers:
\begin{equation}
    \label{}
    \hat{T}_{\rm{QDM}} = \sum_{\bplqt} 
            \left|
                \vrtplqt
            \right\rangle
            \left\langle
                \hrzplqt
            \right|
            +
            h.c..
\end{equation}

Because of the local constraints, all states in the Hilbert space of a GQDM are invariant under local gauge transformations a the vertices, $G_v$~\cite{Moessner2001}. For GQDM with a number constraint, we write the gauge transformation as:
\begin{equation}
G_v = e^{i \alpha \left( n_v -n_0 \right)}; \notag
\end{equation}
as the states in the GQDM Hilbert space are invariant under $G_v$ for all $0 \leq \alpha \leq 2 \pi$, this represents a local $U(1)$ gauge symmetry. When the number constraint is relaxed to a parity constraint, there is a correspondingly reduced gauge symmetry which can be written as
\begin{equation}
G_v^e = e^{\pm i \pi n_v}, \quad G_v^o = e^{\pm i \pi \left(n_v-1\right)}, \notag
\end{equation}
where $G^{e\left(o\right)}_v$ is the transformation for the even (odd) parity constraint; the gauge symmetry is now reduced to to $Z_2$.
By construction, the Hamiltonian is gauge invariant and thus commutes with $G_v$
\begin{equation}
\left[ H_{\rm{GQDM}}, G_v \right] = 0. \notag
\end{equation}

>>>>

\newcommand{\ket}[1]{| #1 \rangle}
\newcommand{\eket}[1]{\left | #1 \right \rangle}
\newcommand{\ebra}[1]{\left \langle #1 \right |}
\newcommand{\Eqref}[1]{Eq. \eqref{#1}}
\newcommand{\RK}{\mathrm{RK}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

@@

\begin{figure}[t!]
    \centering
    \includegraphics[width=0.5\columnwidth]{QDPM_ex_config.pdf}
    \caption{Example of an allowed configuration in the quantum dimer-pentamer model.}
    \label{fig:QDPMex}
\end{figure}

@@

\begin{figure}[htpb]
    \centering
    \includegraphics[width=0.8\linewidth]{example_local_constraints.pdf}
    \caption{Clockwise from the upper left panel we show an example of a QDM with constraints: one
    dimer ($n_0 = 1$), two dimers ($n_0 = 2$), and an even ($n_P = 0$) or odd ($n_P = 1$) number of dimers touching each vertex.}
    \label{fig:example_local_constraints}
\end{figure}

Here we will provide background about generalized quantum dimer models on the square lattice. We consider systems with Ising-like dimer degrees of freedom on each link, such that each link has either $0$ or $1$ dimers. The Hilbert space consists of configurations of dimers on the square lattice that satisfy a local constraint at each vertex. Defining $n_v$ as the number of dimers that touch a vertex, we can distinguish between fixed number constraints of the form $n_v = n_0$ and fixed parity constraints of the form $n_v\mod 2 = n_P$. The unique, non-trivial constraints of this form for the square lattice are the number constraints $n_0=1,2$ and the parity constraints $n_P =0,1$ (see \figref{fig:example_local_constraints}).

The Hamiltonian of a GQDM is of the form
\begin{equation}
H_{\rm{GQDM}} = -\sum_\nu t_\nu \hat{T}_{\nu} + \sum_\nu v_\nu \hat{V}_{\nu} \label{eq:HGQDM}
\end{equation}
where $\hat{T}_\nu$ are off-diagonal kinetic energy operators, $\hat{V}_\nu$ are diagonal the potential energy operators, $t_\nu$ and $v_\nu$ parametrize the strength of these terms, and each $\nu$ and represents a set of $n_\nu$ neighboring links. $\hat{T}_\nu$ is constructed to favor a resonance between two distinct allowed dimerazations of $\nu$, $c_\nu$ and its compliment $\bar{c}_\nu$, and is written as:
\begin{equation}
\hat{T}_{\nu} = \eket{c_\nu}\ebra{\bar{c}_\nu} + h.c.;
\end{equation}
in the simplest case, $\hat{T_\nu}$ involves the minimum non-trivial dimer dynamics that preserve the constraint. Similarly, the potential energy term provides an energy cost for states with these local configurations:
\begin{equation}
\hat{V}_{\nu} = \eket{c_\nu}\ebra{c_\nu} +  \eket{\bar{c}_\nu}\ebra{\bar{c}_\nu}.
\end{equation}
In general the $T_\nu$'s don't commute with each other or with the $V_\nu$'s, and thus the ground state of $H_{\rm{GQDM}}$ is not explicitly known.

However, at the special point where $t_\nu = v_\nu~\forall \nu$, $H_{\rm{GQDM}}$, known as the Rokhsar-Kivelson (RK) point, the exact ground state is known. This can be shown by writing \Eqref{eq:HGQDM} at the RK point as the sum of projection operators
\begin{equation}
H_{\rm{RK}} = 2 \sum_\nu \hat{h}_\nu \label{eq:HRK},
\end{equation}
where $\hat{h}_\nu \equiv \hat{P}^\dagger_{\nu} \hat{P}_{\nu}$ and $\hat{P}^\dagger_\nu$ is a vector orthogonal to the equal superposition of $c_\nu$ and $\bar{c}_\nu$
\begin{equation}
\hat{P}^\dagger_\nu \equiv \frac{1}{\sqrt{2}} \biggl( \eket{ c_\nu} - \eket{\bar{c}_\nu} \biggr).
\end{equation} 
Since all $h_\nu$ are projectors with eigenvalues $\{0,1\}$, a zero energy ground state of \Eqref{eq:HRK} is simultaneously annihilated by all $\hat{h}_\nu$. Such a zero energy state, known as the RK state, may be formed from an equal superposition of all dimerizations that are connected by $\{T_\nu\}$:
\begin{equation}
\eket{\Phi_{\RK}^{\Omega}} = \sum_{C \in \Omega} \eket{C}
\end{equation} 
where $\Omega$ is a subset of the Hilbert space connected by $\{T_\nu\}$. Thus $\{ \ket{\Phi_{\RK}^\Omega} \}$ are the exact, degenerate ground states of $H_{\rm{GQDM}}$ with a degeneracy labelled by $\Omega$ at the RK point.

Because of the local constraints, all states in the Hilbert space of a GQDM are invariant under local gauge transformations a the vertices, $G_v$~\cite{Moessner2001}. For GQDM with a number constraint, we write the gauge transformation as:
\begin{equation}
G_v = e^{i \alpha \left( n_v -n_0 \right)}; \notag
\end{equation}
as the states in the GQDM Hilbert space are invariant under $G_v$ for all $0 \leq \alpha \leq 2 \pi$, this represents a local $U(1)$ gauge symmetry. When the number constraint is relaxed to a parity constraint, there is a correspondingly reduced gauge symmetry which can be written as
\begin{equation}
G_v^P = e^{\pm i \pi \left(n_v-n_P\right)}, \notag
\end{equation}
and the gauge symmetry is now reduced to to $Z_2$.
By construction, the Hamiltonian is gauge invariant and thus commutes with $G_v$
\begin{equation}
\left[ H_{\rm{GQDM}}, G_v \right] = 0. \notag
\end{equation}

Given the gauge symmetry, one may construct a gauge invariant string operator along an oriented path $\Gamma$ path on the dual lattice
\begin{equation}
S_\Gamma \equiv \prod_{\ell \in \Gamma} e^{i \alpha_\ell n_\ell}
\end{equation}
where $\alpha_\ell$ is a phase that is associated with each link on the lattice, as well as the direction the link is crossed in $\Gamma$.

\subsection{Square lattice QDM}

the simplest constraint is that of the hard-core fully-packed QDM where a single dimer must touch each vertex. We write this constraint as $n_v = 1$,

For example in the QDM $\hat{T}$ is the sum of all possible ``plaquette flip" terms acting on plaquettes with parallel dimers:
\begin{equation}
    \label{}
    \hat{T}_{\rm{QDM}} = \sum_{\bplqt} 
            \left|
                \vrtplqt
            \right\rangle
            \left\langle
                \hrzplqt
            \right|
            +
            h.c..
\end{equation}
